Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-11-27 GroupPolarizationStatmod/paper/ms.tex
added GP from simple consensus fig, model text

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\linenumbers
\modulolinenumbers[5]

@@

\begin{itemize}
  \item 
    \emph{Group polarization} is a process where the mean opinion of a 
    like-minded group becomes more extreme over time~\cite{Brown1986,Brown2000}. 
    It is therefore an important component process in
    the broader process of increasing political polarization that can hamstring
    governments~\cite{Mason2018UncivilAgreementBook,Klein2020}.
  \item
    Individuals' opinions in group polarization experiments are typically 
    measured using an ordinal scale, such
    as a Likert scale, where behavioral experiment participants' internal, theoretically
    continuous, opinions are converted to one of several opinion bins indiciating
    their degree of agreement or disagreement with some prompt statement.
  \item
    Recently, however, it has been shown that it is critical to use 
    appropriate statistical methods when using experimental designs that 
    measure continuous latent psychological variables with ordinal
    measurements---failure to do so can result in false positive support for
    the presence of experimental effects~\cite{Liddell2018}.
  \item
    In this paper, we develop a generative modeling approach for evaluating whether or
    not published detections of group polarization are plausibly false, i.e.,
    whether they are \emph{plausibly-false positives} (PFPs). 
  \item
    We found significant evidence that many published group polarization findings
    are indeed plausibly false since we found distribution parameters that
    generated simulated pre- and post-discussion data that a $t$-test found
    to be from distributions with different means, although there was no difference
    in mean, only a difference in variance. 
  \item
    Although a $t$-test may be fooled by simple consensus under ordinal measurements,
    we show that a more appropriate ordered probit model of the data yield no
    such plausibly-false positives.
  \item
    Our work shows the need for using appropriate statistical tests, as well as
    the need for a consistent experimental paradigm so that meaningful meta-analyses
    can be performed for theoretical integration and rigorous understanding of
    the group polarization effect, should it prove to occur in future group
    polarization studies that use proper statistics. In addition to avoiding
    $t$-tests, future group polarization studies can be more reliable by
    accounting for random effects arising from, e.g., socioeconomic status,
    gender, and other individual-level factors that may bias how much and in what
    direction a participant changes their opinions. Failure to do so can result 
    in artificially inflated effect sizes~\cite{Yarkoni2021}.
\end{itemize}

@@

We assume that participant $i$ has latent opinion at time $t$ that is drawn from
a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation}
\noindent 
This distribution has mean $\mu_t$ and standard deviation $\sigma_t$.
This form is chosen to
reflect the common statistical analysis used by group polarization studies,
not because it is statistically appropriate---in fact it is not, which is what
we aim to show in this paper.

@@

(HOW SIMPLE CONSENSUS LOOKS LIKE GROUP POLARIZATION)

@@

We model measurement of $N \to \infty$ participant opinions,
which results in a histogram of frequency of opinions
in each bin, i.e., $o_{i,t} = b_k$. The frequency of responses in each bin is 
the integral over the continuous normal probability density function from 
one bin threshold to another. This transforms the probability density 
function from $p(o_{i,t};~\mu_t, \sigma_t)$ to the probability
of observing a reported opinion of each bin value. 
The probability of observing an opinion in bin
$b_k$ is calculated by integrating the normal probability density 
function over the range of $\theta_{k-1}$ to $\theta_{k}$. 
Formally, we write the 
probability of observing an opinion in bin $k$ at time $t$ as
\begin{equation}
\begin{aligned}
  p(o = b_k;\mu_t, \sigma_t, \theta) 
    &= \int_{\theta_{k-1}}^{\theta_k} p(o;\mu_t, \sigma_t) do \\
    &= \Phi \left( \frac{o - \mu_t}{\sigma_t} \right)\Big|_{o=\theta_{k-1}}^{\theta_k} \\
    &= \Phi \left( \frac{\theta_k - \mu_t}{\sigma_t} \right) - 
       \Phi \left( \frac{\theta_{k-1} - \mu_t}{\sigma_t} \right)
\end{aligned}
\label{eq:binFrequency}
\end{equation}
\noindent
where $\Phi(\frac{o - \mu}{\sigma})$ is the normalized 
normal cumulative distribution function over opinions $o$, shifted by an amount $\mu$ with
standard deviation $\sigma$.
\subsubsection{Identify null-model parameters that yield plausibly-false positives}

@@

\vspace{.5em}
\spacing{1.0}
% \input{plausibleFP-table.tex}
  \begin{table}
    \begin{center}
      \caption{Number and fraction of experimental conditions where group polarization was
               reported to have occurred, but plausibly did not.}
      \vspace{.5em}
      \input{plausibleFP-table-custom.tex}
    \end{center}
  \end{table}
\spacing{1.5}

@@

\spacing{1.0}
\hspace{-1em}
\begin{center}
\input{t-test-table.tex}
\end{center}
% \input{t-test-table-custom.tex}
\spacing{1.5}

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{float}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{caption}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\linenumbers
\modulolinenumbers[3]

@@

\begin{itemize}
  \item 
    \emph{Group polarization} is a process where the mean opinion of a 
    like-minded group becomes more extreme over time~\cite{Brown1986,Brown2000}. 
    It is therefore an important component process in
    the broader process of increasing political polarization that can hamstring
    governments~\cite{Mason2018UncivilAgreementBook,Klein2020}.
  \item
    Individuals' opinions in group polarization experiments are typically 
    measured using an ordinal scale, such
    as a Likert scale, where behavioral experiment participants' internal, theoretically
    continuous, opinions are converted to one of several opinion bins indiciating
    their degree of agreement or disagreement with some prompt statement.
  \item
    Recently, however, it has been shown that it is critical to use 
    appropriate statistical methods when using experimental designs that 
    measure continuous latent psychological variables with ordinal
    measurements---failure to do so can result in false positive support for
    the presence of experimental effects~\cite{Liddell2018}.
  \item
    In this paper, we develop a generative modeling approach for evaluating whether or
    not published detections of group polarization are plausibly false, i.e.,
    whether they are \emph{plausibly-false positives} (PFPs). 
  \item
    We found significant evidence that many published group polarization findings
    are indeed plausibly false since we found distribution parameters that
    generated simulated pre- and post-discussion data that a $t$-test found
    to be from distributions with different means, although there was no difference
    in mean, only a difference in variance. 
  \item
    Although a $t$-test may be fooled by simple consensus under ordinal measurements,
    we show that a more appropriate ordered probit model of the data yield no
    such plausibly-false positives.
  \item
    Our work shows the need for using appropriate statistical tests, as well as
    the need for a consistent experimental paradigm so that meaningful meta-analyses
    can be performed for theoretical integration and rigorous understanding of
    the group polarization effect, should it prove to occur in future group
    polarization studies that use proper statistics. In addition to avoiding
    $t$-tests, future group polarization studies can be more reliable by
    accounting for random effects arising from, e.g., socioeconomic status,
    gender, and other individual-level factors that may bias how much and in what
    direction a participant changes their opinions. Failure to do so can result 
    in artificially-inflated effect sizes~\cite{Yarkoni2021}.
\end{itemize}

@@

We assume that participant $i$ has latent opinion at time $t$ that is drawn from
a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation}
\noindent 
This distribution has mean $\mu_t$ and standard deviation $\sigma_t$.
This form is chosen to
reflect the common statistical analysis used by group polarization studies,
not because it is statistically appropriate---in fact it is not, which is what
we aim to show in this paper.

@@

\subsubsection{How simple consensus can look like group polarization}

\begin{table}[h]
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $k \in \{0,\ldots,K\} $ & Bin index ($k$) between 0 and total number of
         ordinal bins, $K$ \\
   $\mu$  &   Latent mean, assumed constant in simple consensus \\ 
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma^2$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\theta_k$  & The upper threshold value for binning latent opinions into
   bin $b_k$ \\
   $\langle o_t \rangle$  &   Observed mean opinion at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

\begin{figure}
  \caption{\emph{Simple consensus} occurs in a group when its latent opinion
  distribution variance decreases, but the mean does not change (A). When 
latent opinions are measured on an ordinal scale, the mean of the distribution
can be distorted by ceiling (or floor) effects (B). Group polarization 
can appear to have occurred where the mean of the observed opinions appears
to increase despite the latent mean being constant (C).}
  \label{fig:consensusDistroIllustration}
  \vspace{1em}
  \centering
    \includegraphics[width=1.0\textwidth]{ConsensusDistroIllustration/figure.pdf}
\end{figure}

@@

We model measurement of $N \to \infty$ participant opinions,
which results in a histogram of frequency of opinions
in each bin, i.e., $o_{i,t} = b_k$. The frequency of responses in each bin is 
the integral over the continuous normal probability density function from 
one bin threshold to another. This transforms the probability density 
function from $p(o_{i,t};~\mu_t, \sigma_t)$ to the probability
of observing a reported opinion of each bin value. 
The probability of observing an opinion in bin
$b_k$ is calculated by integrating the normal probability density 
function over the range of $\theta_{k-1}$ to $\theta_{k}$. 
Formally, we write the 
probability of observing an opinion in bin $k$ at time $t$ as
\begin{equation}
\begin{aligned}
  p(o \in b_k; \mu_t, \sigma_t, \theta) 
    &= \int_{\theta_{k-1}}^{\theta_k} p(o;\mu_t, \sigma_t) do \\
    &= \Phi \left( \frac{o - \mu_t}{\sigma_t} \right)\Big|_{o=\theta_{k-1}}^{\theta_k} \\
    &= \Phi \left( \frac{\theta_k - \mu_t}{\sigma_t} \right) - 
       \Phi \left( \frac{\theta_{k-1} - \mu_t}{\sigma_t} \right)
\end{aligned}
\label{eq:binFrequency}
\end{equation}
\noindent
where $\Phi(\frac{o - \mu}{\sigma})$ is the normalized 
normal cumulative distribution function over opinions $o$, shifted by an amount $\mu$ with
standard deviation $\sigma$.
\subsubsection{Identify null-model parameters that yield plausibly-false positives}

@@

\vspace{.5em}
\spacing{1.0}
% \input{plausibleFP-table.tex}
\begin{table}[H]
    \begin{center}
      \caption{Number and fraction of experimental conditions where group polarization was
               reported to have occurred, but plausibly did not.}
      \vspace{.5em}
      \input{plausibleFP-table-custom.tex}
    \end{center}
  \end{table}
\spacing{1.5}

@@

\spacing{1.0}
\hspace{-1em}
\input{t-test-table.tex}
% \input{t-test-table-custom.tex}
\spacing{1.5}
